Data Specialization for a Staged Lambda Calculus

### 2014-11-11 paper/example.tex

\ur{
\begin{itemize}
\item 
I find it hard to distinguish between blue and black on printed paper.
\item
$\pause$ should be called ``promote'' or something like that. I
  believe this is what Taha calls cross stage persistence.  THat work should
  be cited.
\item the splitting example should shed more light to the techniques
  and try to give more insight.
\end{itemize}
}

\crem{
Outline:
\begin{enumerate}
\item Say we want to run a bunch of quickselect queries on a single list.
\item We observe that some of the code depends only on the list.
\item We can write this code in \lang\ to make that dependency clear.
\item We can intuitively split into two functions.
\item This is asymptotically better.
\end{enumerate}
}

@@

Staging is a technique for using language constructs and types to
explicitly divide a computation according to the frequency or the
arrival time of data.

@@

Quickselect, which we define in \ref{fig:qs-unstaged}, first partitions the list
by using the first element as a pivot%
\footnote{We assume that the list is prepermuted to guarantee the expected
linear time behavior.}
and then recurs on one of the two resulting sides to find the desired element. The
side chosen is determined by the relationship of \texttt{k} to the size of the
first half \texttt{n}, which is returned by the \texttt{partition} function
along with the two sides.

Now, consider an application where we perform many order statistics queries
on the same collection \texttt{l}, but with $m$ different ranks \texttt{k}:
%
\begin{lstlisting}
quickselect l 1
quickselect l 2
 @$\ldots$@ 
quickselect l @$m$@.
\end{lstlisting}
%
To take advantage of this, we can \emph{stage} quickselect so that the parts of
the computation that depend on the collection~\texttt{l} is performed once, and
the desired result is computed by using~\texttt{k} and the results from this
prior stage.

%% There has been the extensive research into the question of how to
%% automatically add staging annotations to unstaged code.  This process
%% is known as {\em binding time analysis}, and we do not consider it
%% here.  Instead, we assume that all input programs are properly staged
%% according to some programmer's intent.

\crem{missing text}

Using techniques essentially the same as
prior work, we can stage our quickselect code by using staged typing
and inserting three keywords~\texttt{next, prev, hold} into the
appropriate places in the code as shown in \ref{fig:qs-staged}. To
write the staged version, the programmer first observes that the
rank~\texttt{k} has now the type $\fut\rmint$ (read ``circle int'')
rather than $\rmint$, indicating that it will be available in the
second stage.  All types that are not of circle type are available in
the first stage.  Naturally the output type, representing the $k$th
largest element of the list, is also $\fut\rmint$, since it is not
computed until the second stage.

\crem{explain circle types}

The construct $\next$ occurs in a stage \bbone\ context and indicates that the
contents of its block are stage \bbtwo, whereas $\prev$ occurs in a stage
\bbtwo\ context and indicates that the contents of its block is stage \bbone.
(We adopt the convention that the top-level context is stage \bbone.)

Informally speaking, the $\next$ construct enables operating on second stage
computations in the first stage essentially by treating them as boxes that can
be moved around, stored in data structures, and passed to functions, and the
$\prev$ construct enables unwrapping the contents of such a box to use in the
second stage.

For convenience, the code also uses $\pause$ blocks.  This construct
has same stage signature as $\prev$, but instead of ``unwrapping''
$\fut$ types it simply promotes integers from stage~\bbone\ to
stage~\bbtwo.%
\footnote{$\pause$ can be implemented from our other language features, with
some effort. We provide it as a primitive to shorten examples. Furthermore, it
would be wise to extend $\pause$ to other base types, if we had them, and to
products and sums thereof.  This is related to the notion of {\em mobility} in
\cite{murphy05} and {\em stability} in \cite{krishnaswami13}.}

We note that given the type signature of staged quickselect, there are several
ways to insert the staging annotations.  The version we presented here maximizes
the work done at stage~\bbone.

The type system guarantees that all first-stage computations depend only on the
first stage values and can indeed be computed in the absence of second-stage
values.

\crem{partial evaluation requires compile-time access to first stage, whereas we
don't}

@@

Here, we wish to derive two functions, \texttt{quickselect1} and
\texttt{quickselect2}, that are semantically equivalent to quickselect in
the sense that:
\\
\texttt{quickselect l k = quickselect2 (quickselect1 l) k}.
\\
\crem{and moreover they correspond to the stages}

Since \texttt{quickselect1} requires only on a first stage
argument~\texttt{l}, we can run it once to produce a boundary data
structure that can be used for all subsequent queries on~\texttt{l},
e.g.,  the code consisting of $m$ calls 
\begin{lstlisting}
quickselect l 1
quickselect l 2
 @$\ldots$@ 
quickselect l @$m$@
\end{lstlisting}
%
can be replaced by 
%
\begin{lstlisting}
b = quickselect1 l
quickselect2 b 1
quickselect b 2
@$\ldots$@ 
quickselect b @$m$@.
\end{lstlisting}

In this paper, we show that it is possible to generalize data
specialization to a fully general language with recursion and
recursive types.  The key ideas behind our approach are: ...

\crem{walk through manual example here}

The final version, cleaned up and given better datatype names, is shown in
\ref{fig:qssplit}.  Observe that \texttt{qSelect1} essentially builds a binary
search tree from the input list, and \texttt{qSelect2} searches that tree for
the correct index.

@@

Note that the desired optimized code shows above is intellectually more
sophisticated than the code that we have started with: the optimized code is
able to create a data structure, a balanced binary tree augmented with indexing
information, and use a binary search technique over this tree to compute the
result asymptotically more efficiently.
%In fact, based our teaching experience, we can imagine this kind of problem to
%be a moderately difficult exam question in an undergraduate algorithms class, as
%it not only requires understanding of data structures such as binary search
%trees but also requires modifying them to augment with indexing information to
%support rank-based search.

This paper formally describes and analyzes the splitting transformation
exemplified above.  (I also want to make it clear that recognizing \lang's
appropriateness for this is itself is a contribution.)

>>>>

\ur{
\begin{itemize}
\item 
I find it hard to distinguish between blue and black on printed paper.
\item
$\pause$ should be called ``promote'' or something like that. I
  believe this is what Taha calls cross stage persistence.  THat work should
  be cited.
\end{itemize}
}

@@

@@

Quickselect, which we define in \ref{fig:qs-unstaged}, first partitions the list
by using the first element as a pivot%
\footnote{We assume that the list is prepermuted to guarantee the expected
linear time behavior.}
and then recurs on one of the two resulting sides to find the desired element.
The side chosen is determined by the relationship of \texttt{k} to the size of
the first half \texttt{n}, which is returned by the \texttt{partition} function
along with the two sides themselves.

Now, consider an application where we perform many order statistics queries on
the same collection \texttt{l}, but with $m$ different ranks \texttt{k}.
Certainly, it is possible to implement this with $m$ calls to
\texttt{quickselect}:
%
\begin{lstlisting}
quickselect l @$k_1$@
quickselect l @$k_2$@
 @$\vdots$@ 
quickselect l @$k_m$@.
\end{lstlisting}
%
Can we do better?

\subsection{Staging}

An astute programmer might notice that much of the code in \texttt{quickselect}
does \emph{not} depend on the rank \texttt{k}---for example, the list is
partitioned before \texttt{k} is ever used. Moreover, while the recursive calls
to \texttt{qSelect} are guarded by a comparison to \texttt{k}, the lists passed
in to those recursive calls also don't depend on \texttt{k}. Thus, if we are
willing to change the evaluation order of the language (and evaluate under the
\texttt{case} statement), it should be possible to set things up so that
\emph{all} the computation involving \texttt{l} is performed before \emph{any}
of the computation involving \texttt{k}.

We can formalize this intuition by writing quickselect in \lang\
(\ref{sec:semantics}), a \emph{staged}, typed lambda calculus. In \lang, we can
directly express the idea that the argument \texttt{l} is known at the
\emph{first} stage of the computation, while the argument \texttt{k} is only
known at the \emph{second}, and all purely-stage-\bbone\ computation occurs
before any stage \bbtwo\ computation.

A staged version of quickselect is written in \ref{fig:qs-staged}, where stage
\bbone\ code is red, and stage \bbtwo\ code is blue. Whereas before
\texttt{qSelect} took a \textrm{list} and \rmint\ and returned an \rmint, now we
must be more careful---at stage \bbone, \texttt{qsStaged} takes a \textrm{list}
and a $\fut\rmint$ (an $\rmint$ \emph{in the future}), and returns a
$\fut\rmint$.

At stage \bbone, all terms of non-circle type are available for
immediate use. Indeed, we are able to case on \texttt{l} as usual, and in the
\texttt{Cons} branch, pass it to the stage \bbone\ function \texttt{partition}.

\crem{Very rough text for the remainder of the section. Be a bit more explicit.}

However, \texttt{k} has type $\fut\rmint$, so we cannot use it as part of a
stage \bbone\ computation. Luckily, we are trying to produce a $\fut\rmint$. To
create one of these, we can wrap $\next$ around a stage \bbtwo\ computation of
an integer; in that computation, we can wrap $\prev$ around the $\fut\rmint$
\texttt{k} in order to use it as a stage \bbtwo\ integer.

(There's a subtle difference here between an \emph{int later} at the first
stage, and an int at the second stage.)

Essentially, $\next$ allows us to pass around opaque stage \bbtwo\ computations
at stage \bbone, and the $\prev$ construct lets us unwrap those computations for
use at stage \bbtwo.

For convenience, the code also uses $\pause$ blocks. This construct has same
stage signature as $\prev$, but instead of ``unwrapping'' $\fut$ types it simply
promotes integers from stage \bbone\ to stage \bbtwo.%
\footnote{$\pause$ can be implemented from our other language features, with
some effort. We provide it as a primitive to shorten examples. Furthermore, it
would be wise to extend $\pause$ to other base types, if we had them, and to
products and sums thereof.  This is related to the notion of {\em mobility} in
\cite{murphy05} and {\em stability} in \cite{krishnaswami13}.}

The process of adding staging annotations to unstaged code has been the subject
of extensive research under the name of \emph{binding time analysis}. In this
paper, we assume that these annotations have been provided by the programmer (or
perhaps a binding time analysis tool), and do not consider the problem of
generating such annotations. There are many ways to annotate any program,
including \texttt{qselect}; the version presented here matches the intuition for
the maximal amount of work that can be performed at stage \bbone.

The purpose of writing quickselect in \lang\ is that, because \texttt{qsStaged}
typechecks, it is a valid way of dividing the program into two stages, where the
stage \bbone\ computations can all be performed prior to the stage \bbtwo\
computations. In \ref{sec:semantics} we provide an evaluation strategy for
\lang\ which validates that claim.

%The construct $\next$ occurs in a stage \bbone\ context and indicates that the
%contents of its block are stage \bbtwo, whereas $\prev$ occurs in a stage
%\bbtwo\ context and indicates that the contents of its block is stage \bbone.

%(We adopt the convention that the top-level context is stage \bbone.)

\subsection{Splitting Staged Programs}

Because we have found a valid staging \texttt{qsStaged}, one might ask what
happens if we split this function into \emph{two} functions, one of which
performs all the stage \bbone\ computations given \texttt{l}, and the second
which performs all the stage \bbtwo\ computations given \texttt{k} and the
results of the previous.

Intuitively, what should this be? All calls to \texttt{partition} are in stage
\bbone, so this means we build a binary search tree from \texttt{l}. All
decisions on which side to recur down happen in stage \bbtwo, so this means we
traverse the tree once we have the rank \texttt{k}.

(There are two kinds of things we must pass to the second function---which stage
\bbtwo\ computations to use, and also anything that was held.)

@@

Formally, we wish to derive two functions, \texttt{qSelect1} and
\texttt{qSelect2}, which respectively contain all the stage \bbone\ and stage
\bbtwo\ content of \texttt{qSelect}, and are moreover semantically equivalent to
\texttt{qSelect} in the sense that:
\\
\texttt{qSelect l k = qSelect2 (qSelect1 l) k}.
\\

Since \texttt{quickselect1} requires only on a first stage
argument \texttt{l}, we can run it once to produce a boundary data
structure that can be used for all subsequent queries on \texttt{l},
e.g., we can perform $m$ queries by simply running:
%
\begin{lstlisting}
let b = qSelect1 l in
  qSelect2 b @$k_1$@
  qSelect2 b @$k_2$@
   @$\vdots$@ 
  qSelect2 b @$k_m$@.
\end{lstlisting}

Indeed, in \ref{sec:splitting,sec:implementation} we describe an algorithm for
splitting staged terms, which on \texttt{qsStaged} actually emits the algorithm
in \ref{fig:qssplit}.

@@

%Note that the desired optimized code shows above is intellectually more
%sophisticated than the code that we have started with: the optimized code is
%able to create a data structure, a balanced binary tree augmented with indexing
%information, and use a binary search technique over this tree to compute the
%result asymptotically more efficiently.
%In fact, based our teaching experience, we can imagine this kind of problem to
%be a moderately difficult exam question in an undergraduate algorithms class, as
%it not only requires understanding of data structures such as binary search
%trees but also requires modifying them to augment with indexing information to
%support rank-based search.

%(I also want to make it clear that recognizing \lang's appropriateness for this
%is itself is a contribution.)
